Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-21 vertical-thermal-review.Rmd
Eleinis' edit included

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From Anderson-Teixeira et al. (in revision). In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020. ](tree_rings/Figure_4_tree_rings.png)

@@

Leaf level $T_{opt}$ is closely related to ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and $g_s$ limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperate and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. 
Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
A daytime temperature threshold of 32$^\circ$C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].

>>>>

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From Anderson-Teixeira et al. (in revision). In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020. ](tree_rings/Figure_4_tree_rings.png)

@@

Leaf level $T_{opt}$ is closely related to the $T_{opt}$ of ecosystem photosynthesis ($T_{optE}$). Temperature limitations to leaf processes, such as VPD and $g_s$ limitation, can modify $T_{optE}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperate and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. 
Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
A daytime temperature threshold of 32$^\circ$C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
